---
title: "PA1_template"
author: "Nick"
date: "May 14, 2015"
output:
  html_document:
    keep_md: yes
    toc: yes
---
Reproducible Research Coursera Course - Programming Assignment 1

The following document is an R Markdown document containing the text and R code for the first programming assignment in the Reproducible Research Coursera Course.  

#Loading and Preprocessing the data  
First, I set the working directory to the desktop which contains the data file
```{r echo=TRUE}
setwd("~/Desktop")
```

Second, I read in the data into a variable called data
```{r}
data <- read.csv("activity.csv")
```

Third, I convert the date variable from a factor to a Date date type
```{r}
str(data)
data$date <- strptime(data$date, "%Y-%m-%d")
data$date <- as.Date(data$date)
str(data)
```

#What is mean total number of steps taken per day?
I calculate the mean 2 seperate ways:  
1. I show the total number of steps taken per day and rename the columns while taking out NA values. Note that this method includes 0 values (total observations = 61)
```{r}
steps_taken_per_day <- aggregate(data$steps, by=list(data$date), FUN=sum, na.rm=TRUE)
colnames(steps_taken_per_day) <- c("Day", "Steps")
```
2. Same as above except this method does not include 0 values (total observations = 53)
```{r}
steps_taken_per_day2 = aggregate(steps~date, data=data, FUN=sum, na.rm=TRUE)
colnames(steps_taken_per_day2) <- c("Day", "Steps")
```

Next, I plot a histogram for both methods above showing the total number of steps taken each day
```{r}
hist(steps_taken_per_day$Steps, main="Total number of steps taken each day (including 0 values)", xlab="Day", breaks = 15)
hist(steps_taken_per_day2$Steps, main="Total number of steps taken each day (not including 0 values)", xlab="Day", breaks = 15)
```

Finally, I calculate the mean and median using both methods  
Method 1:
```{r}
mean(steps_taken_per_day$Steps)
median(steps_taken_per_day$Steps)
```
Method 2:
```{r}
mean(steps_taken_per_day2$Steps)
median(steps_taken_per_day2$Steps)
```
#What is the Average Daily Activity Pattern?
Create variable that aggregates steps by interval while still ignoring NA values
```{r}
interval_steps <- aggregate(data$steps, by=list(data$interval), FUN=mean, na.rm=TRUE)
colnames(interval_steps) <- c("Interval", "Steps")
```

Next plot a time series plot 
```{r}
plot(interval_steps[,1], interval_steps[,2], xlab="5 Minute Interval", ylab="Average Steps per day", main="Total number of steps taken each day", type="l")
```

Finally, show which 5 minute interval on average contains the highest number of steps
```{r}
interval_steps[which.max(interval_steps[,2]),]
```
#Imputting Missing Values
Calculate the total number of missing values in the dataset. The output shows there are 2,304 missing values in the steps variable
```{r}
summary(data)
```

Create a copy of the data file that will contain filled in NA values
```{r}
data2 <- data
```

Missing values are filled in by using the average interval value for that day
```{r}
for(i in 1:nrow(data2)) {
  if(is.na(data2$steps[i])) {
    data2$steps[i] = subset(interval_steps, Interval == data2$interval[i])$Steps
  }
}
summary(data2)  ## Note that there are no NA values
```

I show the total number of steps taken per day (using the data set with missing values filled in) and rename the columns while taking out NA values
```{r}
steps_taken_per_day_complete <- aggregate(data2$steps, by=list(data2$date), FUN=sum, na.rm=TRUE)
colnames(steps_taken_per_day_complete) <- c("Day", "Steps")
```

Next, the histogram shows the total number of steps taken each day
```{r}
hist(steps_taken_per_day_complete$Steps, main="Total number of steps taken each day", xlab="Day", breaks = 15)
```

Lastly, I calculate the mean and median of the data set (note that the output shows that the mean and median are similar to method 2 above which excluded the 0 values)
```{r}
mean(steps_taken_per_day_complete$Steps)
median(steps_taken_per_day_complete$Steps)
```

#Are there any Differences in Activity Patterns between Weekdays and Weekends?
Create and add a column containing the day of week for that particular day
```{r}
weekdays <- weekdays(data2$date)
data2$weekday <- weekdays
```

Create and add a column saying if day is a weekday or a weekend
```{r}
for(i in 1:nrow(data2)) {
  if(data2$weekday[i] == "Saturday" | data2$weekday[i] == "Sunday") {
    data2$daytype[i] <- c("Weekend")
  } else {
    data2$daytype[i] <- c("Weekday")
  }
}
data2$daytype <- as.factor(data2$daytype)
str(data2)
```

Create a time series panel plot of the 5 minute interval and average number of steps taken across all 5 minute intervals for weekdays and weekends
```{r}
interval_steps_complete <- aggregate(data2$steps, by=list(data2$interval, data2$daytype), FUN = mean)
colnames(interval_steps_complete) <- c("interval", "daytype", "steps")

library(lattice)
xyplot(interval~steps|daytype, data=interval_steps_complete, main = "Average across all 5 minute intervals for weekdays and weekends", layout = c(1, 2), xlab = "Interval", ylab="Average Steps")
```
